---
title: "UFO Sightings"
author: "Kit Applegate"
date: "2023-06-20"
output: github_document
---

This Data is from the [National UFO Reporting Center](https://nuforc.org/webreports/ndxshape.html) , cleaned and enriched with data from sunrise-sunset.org by Jon Harmon. IT was part of the this week #tidytuesday data set. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r cars, echo=TRUE, message=FALSE, warning=FALSE}
# Load required libraries

# The 'tidyverse' library is a collection of packages for data manipulation and visualization.
library(tidyverse)

# The 'lubridate' library provides functions for working with dates and times.
library(lubridate)

# The 'chron' library provides functions for handling dates and times in a compact format.
library(chron)

# The 'scales' library provides functions for scaling and formatting data.
library(scales)

# The 'viridis' library provides color scales for visualizations.
library(viridis)

# Read data files

# Read the 'ufo_sightings.csv' file from the specified URL and store it in the 'ufo_sightings' variable.
ufo_sightings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-06-20/ufo_sightings.csv')

# Read the 'shape.List.csv' file from the local directory and store it in the 'shape.List' variable.
shape.List <- read.csv("raw.Data/shape.List.csv")

```



```{r pressure, echo=TRUE}
# Data Manipulation

# Join the 'ufo_sightings' and 'shape.List' dataframes based on the 'shape' and 'Original' columns,
# and store the result in the 'sightings' variable.
sightings <- left_join(ufo_sightings, shape.List %>% unique(), by = c("shape" = "Original")) %>%
  mutate(Shape = coalesce(New, shape)) %>%  # Create a new 'Shape' column by filling missing values with the original 'shape'
  select(-New)  # Remove the 'New' column from the dataframe

# Add a 'Time' column to the 'sightings' dataframe, derived from the 'reported_date_time' column.
sightings$Time <- format(as.POSIXct(sightings$reported_date_time), format = "%H:%M:%S")

# Convert the 'reported_date_time' column to a 'Date' format and store it in the 'Date' column.
sightings$Date <- as.Date(as.POSIXct(sightings$reported_date_time, 'GMT'))

# Add 'Year', 'Month', and 'Day' columns to the 'sightings' dataframe using the 'Date' column.
sightings <- sightings %>%
  dplyr::mutate(
    Year = lubridate::year(Date),
    Month = lubridate::month(Date),
    Day = lubridate::day(Date)
  )

# Filter the 'sightings' dataframe to include only records with 'country_code' equal to "US",
# and select specific columns for analysis.
sightings <- sightings %>%
  filter(country_code == "US") %>%
  select(
    "Year", "Month", "Day", "Time", "city", "state", "shape", "duration_seconds",
    "summary", "day_part"
  )

# Convert the 'Month' column to a factor with ordered levels.
sightings$Month <- factor(month.abb[sightings$Month], levels = month.abb)

# Filter the 'sightings' dataframe to remove records with "HOAX" in the 'summary' column (case-insensitive).
sightings_filtered <- sightings %>%
  filter(!grepl("HOAX", summary, ignore.case = TRUE))

# Perform additional mutations on the 'sightings_filtered' dataframe, including calculations and recoding of columns.
sightings.Mutate <- sightings_filtered %>%
  mutate(
    minutes = round(duration_seconds / 60, 2),
    hours = round(minutes / 60, 2),
    days = round(hours / 24, 2),
    shape = ifelse(shape == "unknown" | is.na(shape), "other", shape),
    day_part = case_when(
      day_part %in% c("nautical dusk", "astronomical dusk", "civil dusk") ~ "dusk",
      day_part %in% c("civil dawn", "nautical dawn", "astronomical dawn") ~ "dawn",
      TRUE ~ day_part
    )
  )

# Convert the 'Time' column in the 'sightings.Mutate' dataframe to the 'times' class.
sightings.Mutate$Time <- times(format(sightings.Mutate$Time, format = "%H:%M:%S"))

# Convert the 'Month' column in the 'sightings.Mutate' dataframe to a factor.
sightings.Mutate$Month <- as.factor(sightings.Mutate$Month)


```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Data Visualization

# Group the 'sightings.Mutate' dataframe by 'Month' and 'day_part', and calculate the total number of sightings in each group.
# Store the result in the 'total' column and arrange the dataframe in descending order of 'total'.
sightings_summary <- sightings.Mutate %>%
  group_by(Month, day_part) %>%
  summarise(total = n()) %>%
  arrange(desc(total))

# Remove any rows with missing values (NA) from the 'sightings_summary' dataframe.
sightings_summary <- na.omit(sightings_summary)

# Create a bar plot using 'ggplot', mapping 'Month' to the x-axis, 'total' to the y-axis, and 'day_part' to fill and color.
# Use 'geom_col()' to display the bars, and set the plot labels and title.
# Customize the fill and color scales using the 'rainbow' color palette.
ggplot(sightings_summary, aes(x = Month, y = total, fill = day_part, color = day_part)) +
  geom_col() +
  labs(title = "UFO Sightings by Month",
       subtitle = "Part of the Day",
       y = "Frequency",
       x = "") +
  scale_fill_manual(values = rainbow(length(unique(sightings.Mutate$day_part)))) +
  scale_color_manual(values = rainbow(length(unique(sightings.Mutate$day_part))))

```

The provided bar graph illustrates the frequency of sightings categorized by month, accompanied by the corresponding time of day during which these sightings took place. Notably, the month of July stands out as the period with the highest number of reported sightings, particularly during the dusk hours. Based on this analysis, it is reasonable to hypothesize that the observed pattern could potentially be attributed to the presence of fireworks, thereby suggesting a plausible causal relationship between fireworks and the reported sightings.

```{r echo=TRUE, message=FALSE, warning=FALSE}
sighting_times <- strptime(sightings.Mutate$Time, format = "%H:%M:%S")

# Extract the hour component from the POSIXlt format
sighting_hours <- as.integer(format(sighting_times, format = "%H"))

# Create a histogram using the extracted hour component
hist(sighting_hours, breaks = 25, xlab = "Hour", ylab = "Frequency", main = "UFO Sightings Over 24 Hours", col = rainbow(24),
     xlim = c(.5, 24))
```

The depicted histogram showcases the distribution of sightings within a 24-hour timeframe. It is evident that the majority of sightings took place during the late night and early morning hours, specifically after midnight and before dawn.    

```{r echo=TRUE, message=FALSE, warning=FALSE}
heatmap_data <- sightings.Mutate %>%
  group_by(day_part, shape) %>%
  summarise(count = n()) %>%
  na.omit()

# Create the heatmap plot
ggplot(heatmap_data, aes(x = shape, y = day_part, fill = count)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(colours = rainbow(10)) +
  labs(x = "Shape", y = "Time of Day", title = "UFO Sightings Heatmap") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The depicted heat map illustrates the occurrences of various shapes alongside the corresponding time of day during which they were sighted. Notably, the nighttime period stands out as having the highest frequency of shape sightings, with "light" being the most commonly reported shape, closely followed by "other."  




